Simple repression draft

### 2020-06-06 doc/section_02_means.tex
Implemented most of GC's edits plus more of my own

\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple repression, let
us consider what we can learn from theory and experimental measurements on the
average level of gene expression in a population of cells. One experimental
strategy that has been particularly useful (if incomplete since it misses out on
gene expression dynamics) is to measure the fold-change in mean expression. The
fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m (R > 0) \rangle}{\langle m (R = 0) \rangle}
\label{eq:fc_def}
\end{equation}
where angle brackets $\left\langle \cdot \right\rangle$ denote the average over
a population of cells and mean mRNA $\langle m\rangle$ is viewed as a function
of repressor copy number $R$. What this means is that the fold-change in gene
expression is a relative measurement of the effect of the transcriptional 
repressor ($R > 0$) on the gene expression level compared to an unregulated
promoter ($R = 0$). The second equality in~\ref{eq:fc_def} follows from assuming
that the translation efficiency, i.e., the number of proteins translated per
mRNA, is the same in both conditions. In other words, we assume that mean
protein level is proportional to mean mRNA level, and that the proportionality
constant is the same in both conditions and therefore cancels out in the ratio.
This is reasonable since the cells in the two conditions are identical except
for the presence of the transcription factor, and the model assumes that the
transcription factor has no direct effect on translation.

@@

The fold-change is  found as the ratio of mean mRNA with and without repressor
as introduced in Eq.~\ref{eq:fc_def}. Invoking that definition results in
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R = 
\beta\Delta\varepsilon_r - \log (R / N_{NS})$.

@@

\subsubsection{The Three-State Equilibrium Model}
Compared to the previous model, here we fine-grain the repressor unbound state
into two separate states: empty, and RNAP bound as shown in (2) in
Figure~\ref{fig1:means_cartoons}(B). This picture was used in~\cite{Garcia2011a}
as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
generalized to incorporate small-molecule inducers that bind the repressor. 
% Interestingly, the effect of this generalization is
% simply to renormalize $R$ from the total number of repressors to a smaller
% effective number, roughly (though not exactly) the number which are unbound by
% inducers.

The mean mRNA, as derived in the appendix from a similar enumeration of states
and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        },
\end{equation}
where the new variables are $\Delta\varepsilon_P$, the difference in RNAP
binding energy to its specific site (the promoter) relative to an average
nonspecific background site, and the RNAP copy number, $P$. The fold-change
again follows immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
in~\fig{fig1:means_cartoons}(B). Thus far, we see that the two thermodynamic
models, despite making different coarse-graining commitments, result in the same
functional form for the fold-change in mean gene expression.  We now explore how
kinetic models fare when faced with computing the same observable.

@@

The master equation governing this model is identical in form to model 2 above,
namely
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices. The
state vector and promoter transition matrix are now
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
                             p_C(m) \\ p_O(m)\end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound ($R$), empty
($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
increasing dimension by one, the only new feature in $\mathbf{K}$ is the
rate $k_O$, representing the (irreversible) rate of open complex formation from
the closed complex. The initiation matrices are given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to nonequilibrium model 2.

@@

\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
similar to model 2, and mathematically it is, but the interpretation is rather
different. In model 2, we interpreted the third state literally as an RNAP-bound
promoter and modeled initiation of a transcript as triggering a promoter state
change, making the hopefully uncontroversial assumption that an RNAP can only
make one transcript at a time. In contrast, in the present model the promoter
state does \textit{not} change when a transcript is initiated. So we no longer
interpret these states as literally RNAP bound and unbound but instead as
coarse-grained ``active'' and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below when we discuss
Fano factors of models.

@@

\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
intuitive analog to model 1, with just two states, repressor bound or unbound,
and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
the unbound state, single mRNA transcripts are produced as a Poisson process
with some characteristic rate $r$. The current model by contrast produces, at
some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
The burst sizes are assumed to be geometrically distributed with
a mean burst size $b$.

@@

\subsubsection{Summary of Results across models for Fold-Changes in Mean
Expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
expression levels? The key outcome of our analysis is the existence of a master
curve shown in \fig{fig1:means_cartoons}(D) to which the fold-change predictions
of all the models collapse. This master curve is parametrized by only two
effective parameters: $\Delta F_R$, which characterizes the number of repressors
and their binding strength to the DNA, and $\rho$, which characterizes all other
features of the promoter architecture. The key assumption underpinning this
result is that no transcription occurs when a repressor is bound to its
operator. Note, however, that we are agnostic about the molecular mechanism
which achieves this; steric effects are one plausible mechanism, but, for
instance, ``action-at-a-distance'' mediated by kinked DNA due to repressors
bound tens or hundreds of nucleotides upstream of a promoter is plausible as
well.

Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
plausible models of a system and search for their common patterns to tease
out which broad features are and are not important.
In our case, the key feature seems to be the
exclusive nature of repressor and RNAP binding, which allows the parameter
describing the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$. In the
graph-theoretic language which was applied to similar problems
in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the repressor-bound state is
a leaf of the graph describing transitions between promoter states. Arbitrary
nonequilibrium behavior can occur on the rest of the promoter state space, but
it may all be swept up in the effective parameter $\rho$, to which the repressor
makes no contribution.

As suggested in~\cite{Chure2019}, we believe this master curve should generalize
to architectures with multiple repressor binding sites, as long as the
exclusivity of transcription factor binding and transcription initiation is
maintained. The interpretation of $\Delta F_R$ is then of an effective free
energy of all repressor bound states. In an equilibrium picture this is simply
given by the log of the sum of Boltzmann weights of all repressor bound states,
which looks like the log of a partition function of a subsystem. In a
nonequilibrium picture, while we can still mathematically gather terms and give
the resulting collection the label $\Delta F_R$, it is unclear if the physical
interpretation as an effective free energy makes sense. As is well known, free
energies cannot be assigned unambiguously to states out of equilibrium because
the free energy change along a generic path traversing the state space is path
dependent, unlike at equilibrium. A consequence of this is that, out of
equilibrium, $\Delta F_R$ is no longer a simple sum of Boltzmann weights.
Instead it resembles a restricted sum of King-Altman diagrams~\cite{King1956,
Hill1966}. Following the work of Hill~\cite{Hill1989}, it may yet be possible to
interpret this expression as an effective free energy, but this remains unclear
to us. We leave this an open problem for future work.

@@

How do we test this and probe the theory with fold-change measurements? There is
a fundamental limitation in that the master curve is essentially a one-parameter
function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the value of $\rho$,
such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
and/or, RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
precision, inferring the potential microscopic contributions to $\rho$, buried
inside a log no less, from fold-change measurements seems beyond reach. As a
statistical inference problem it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

@@

One might instead imagine a bottom-up modeling approach, where we mathematicize
a model of what we hypothesize the important steps are and are not, use in vitro
data for the steps deemed important, and \textit{predict} what $\rho$ should be.
But again, because of the one-parameter nature of the master curve, many
different models will likely make indistinguishable predictions, and without any
way to experimentally perturb in vivo, there is no clear way to test whether the
modeling assumptions are correct.

@@

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well justified on
the basis of the number of RNAP per cell and the strength of binding of RNAP to
DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
factor? A resolution is offered by an independent inference of $\rho$ in the
absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
(looking ahead, this is equivalent to model 4
in Figure~\ref{fig2:constit_cartoons}(A)),
to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

@@

We would also like to highlight the relevance of~\cite{Landman2019} here.
Landman et.\ al.\ reanalyzed and compared in vivo and in vitro data on the lacI
repressor's binding affinity to its various operator sequences. (The in vivo
data was from, essentially, fitting our master curve to expression
measurements.) They find broad agreement between the in vitro and in vivo
values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$
repressor binding energies do in fact represent real physical free energies.
Again, \textit{a priori} this did not have to be the case, even knowing that
$\rho\approx1$.

>>>>

\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple repression, let
us consider what we can learn from theory and experimental measurements on the
average level of gene expression in a population of cells. One experimental
strategy that has been particularly useful (if incomplete since it misses out on
gene expression dynamics) is to measure the fold-change in mean expression. The
fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m (R > 0) \rangle}{\langle m (R = 0) \rangle}
\label{eq:fc_def}
\end{equation}
where angle brackets $\left\langle \cdot \right\rangle$ denote the average over
a population of cells and mean mRNA $\langle m\rangle$ is viewed as a function
of repressor copy number $R$. What this means is that the fold-change in gene
expression is a relative measurement of the effect of the transcriptional 
repressor ($R > 0$) on the gene expression level compared to an unregulated
promoter ($R = 0$). The second equality in Eq.~\ref{eq:fc_def} follows from assuming
that the translation efficiency, i.e., the number of proteins translated per
mRNA, is the same in both conditions. In other words, we assume that mean
protein level is proportional to mean mRNA level, and that the proportionality
constant is the same in both conditions and therefore cancels out in the ratio.
This is reasonable since the cells in the two conditions are identical except
for the presence of the transcription factor, and the model assumes that the
transcription factor has no direct effect on translation.

@@

The fold-change is  found as the ratio of mean mRNA with and without repressor
as introduced in Eq.~\ref{eq:fc_def}. Invoking that definition results in
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
which matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R = 
\beta\Delta\varepsilon_r - \log (R / N_{NS})$.

@@

\subsubsection{The Three-State Equilibrium Model}
Compared to the previous model, here we fine-grain the repressor unbound state
into two separate states: empty, and RNAP bound as shown in (2) in
Figure~\ref{fig1:means_cartoons}(B). This picture was used in~\cite{Garcia2011a}
as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
generalized to incorporate small-molecule inducers that bind the repressor. 
The effect of this generalization is, roughly speaking,
simply to rescale $R$ from the total number of repressors to a smaller
effective number of available repressors which are unbound by inducers.
We point out that the same generalization can be incorporated quite easily
into any of our models in Figure~\ref{fig1:means_cartoons} by simply
rescaling the repressor copy number $R$ in the equilibrium models, or
equivalently $k_R^+$ in the nonequilibrium models.

The mean mRNA copy number, as derived in Appendix~\ref{sec:non_bursty}
from a similar enumeration of states and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        },
\end{equation}
where the new variables are $\Delta\varepsilon_P$, the difference in RNAP
binding energy to its specific site (the promoter) relative to an average
nonspecific background site, and the RNAP copy number, $P$. The fold-change
again follows immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
in~\fig{fig1:means_cartoons}(B). Thus far, we see that the two thermodynamic
models, despite making different coarse-graining commitments, result in the same
functional form for the fold-change in mean gene expression.  We now explore how
kinetic models fare when faced with computing the same observable.

@@

The master equation governing this model is identical in form to model 2 above,
namely
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\end{equation}
but with a higher-dimensional state space and different matrices. The
state vector and promoter transition matrix are now
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
                             p_C(m) \\ p_O(m)\end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound ($R$), empty
($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
increasing dimension by one, the only new feature in $\mathbf{K}$ is the
rate $k_O$, representing the rate of open complex formation from
the closed complex, which we assume for simplicity to be irreversible
in keeping with some~\cite{Mitarai2015} but not all~\cite{DeHaseth1998}
past literature. The initiation matrices are given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to nonequilibrium model 2.

@@

\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
similar to model 2, and mathematically it is, but the interpretation is rather
different. In model 2, we interpreted the third state literally as an RNAP-bound
promoter and modeled initiation of a transcript as triggering a promoter state
change, making the assumption that an RNAP can only
make one transcript at a time. In contrast, in the present model the promoter
state does \textit{not} change when a transcript is initiated. So we no longer
interpret these states as literally RNAP bound and unbound but instead as
coarse-grained ``active'' and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below when we discuss
Fano factors of models.

@@

\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
intuitive analog to model 1, with just two states, repressor bound or unbound,
and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
the unbound state, single mRNA transcripts are produced as a Poisson process
with some characteristic rate $r$. The current model by contrast produces, at
some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
The burst sizes are assumed to be geometrically distributed with
a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
when we derive this model as a certain limiting case of model 4.

@@

\subsection{Discussion of Results Across Models for Fold-Changes in Mean
Expression}
The key outcome of our analysis of the models in
Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
of all the models collapse. This master curve is parametrized by only two
effective parameters: $\Delta F_R$, which characterizes the number of repressors
and their binding strength to the DNA, and $\rho$, which characterizes all other
features of the promoter architecture. The key assumption underpinning this
result is that no transcription occurs when a repressor is bound to its
operator. Note, however, that we are agnostic about the molecular mechanism
which achieves this; steric effects are one plausible mechanism, but, for
instance, ``action-at-a-distance'' mediated by kinked DNA due to repressors
bound tens or hundreds of nucleotides upstream of a promoter is plausible as
well.

Why does the master curve of Figure~\ref{fig1:means_cartoons}(D) exist at all?
This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
plausible models of a system and search for their common patterns to tease
out which broad features are and are not important.
In our case, the key feature seems to be the
exclusive nature of repressor and RNAP binding, which allows the parameter
describing the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$.
Arbitrary nonequilibrium behavior can occur on the rest of the
promoter state space, but it may all be swept up in the effective
parameter $\rho$, to which the repressor makes no contribution.
We point the interested reader to~\cite{Gunawardena2012}
and~\cite{Ahsendorf2014} for an interesting analysis of similar problems
using a graph-theoretic language.

As suggested in~\cite{Chure2019}, we believe this master curve should generalize
to architectures with multiple repressor binding sites, as long as the
exclusivity of transcription factor binding and transcription initiation is
maintained. The interpretation of $\Delta F_R$ is then of an effective free
energy of all repressor bound states. In an equilibrium picture this is simply
given by the log of the sum of Boltzmann weights of all repressor bound states,
which looks like the log of a partition function of a subsystem. In a
nonequilibrium picture, while we can still mathematically gather terms and give
the resulting collection the label $\Delta F_R$, it is unclear if the physical
interpretation as an effective free energy makes sense. The problem is that free
energies cannot be assigned unambiguously to states out of equilibrium because
the free energy change along a generic path traversing the state space is path
dependent, unlike at equilibrium. A consequence of this is that, out of
equilibrium, $\Delta F_R$ is no longer a simple sum of Boltzmann weights.
Instead it resembles a restricted sum of King-Altman diagrams~\cite{King1956,
Hill1966}. Following the work of Hill~\cite{Hill1989}, it may yet be possible to
interpret this expression as an effective free energy, but this remains unclear
to us. We leave this an open problem for future work.

@@

How do we test this and probe the theory with fold-change measurements? There is
a fundamental limitation in that the master curve is essentially a one-parameter
function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the value of $\rho$,
such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
and/or, RNAP clusters analogous to those observed in
eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
precision, inferring the potential microscopic contributions to $\rho$, buried
inside a log no less, from fold-change measurements seems beyond reach. As a
statistical inference problem it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

@@

One might instead imagine a bottom-up modeling approach, where we mathematicize
a model of what we hypothesize the important steps are and are not, use \textit{in vitro}
data for the steps deemed important, and \textit{predict} what $\rho$ should be.
But again, because of the one-parameter nature of the master curve, many
different models will likely make indistinguishable predictions, and without any
way to experimentally perturb \textit{in vivo}, there is no clear way
to test whether the modeling assumptions are correct.

@@

Another possible route to dissect transcription details without a global
perturbation would be to use phage polymerase with phage-specific promoters.
While such results would carry some caveats, e.g., whether the repression of
the phage polymerase is a good analog to the repression of the native RNAP,
it could nevertheless be worthy of consideration.

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well justified on
the basis of the number of RNAP and $\sigma$ factors per cell
and the strength of binding of RNAP to
DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
factor? A resolution is offered by an independent inference of $\rho$ in the
absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
(looking ahead, this is equivalent to model 4
in Figure~\ref{fig2:constit_cartoons}(A)),
to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

@@

We would also like to highlight the relevance of~\cite{Landman2019} here.
Landman et.\ al.\ reanalyzed and compared \textit{in vivo} and
\textit{in vitro} data on the lacI repressor's binding affinity to its
various operator sequences. (The \textit{in vivo}
data was from, essentially, fitting our master curve to expression
measurements.) They find broad agreement between the
\textit{in vitro} and \textit{in vivo}
values. This reinforces the suspicion that the equilibrium $\Delta\varepsilon_R$
repressor binding energies do in fact represent real physical free energies.
Again, \textit{a priori} this did not have to be the case, even knowing that
$\rho\approx1$.
